The ChEMBL database as Linked Open Data

### 2012-08-18 article.tex
Small edits based on Antony's comments

For drug discovery, the drug-like compounds themselves are the main topic of study.
ChEMBL contains many different compound types, mostly small molecules,
but also peptides, proteins, antibodies, oligosaccharides, oligonucleotides, and
even cells. ChEMBL-RDF follows the approach used by the CHEMINF ontology,
and encodes drugs as classes, rather than instances, and they subclass
classes defined by several ontologies for these types. For example, proteins subclass the
Protein class from the PRotein Ontology (PR\_000000001), small
molecules subclass the Chemical Entity class from the CHEMINF
ontology (CHEMINF\_000000) (which is itself equivalent to the Chemical Entity class in ChEBI, CHEBI\_24431), 
and oligosaccharides and oligonucleotides
subclass their respective matches in the CHEBI ontology (CHEBI\_50699
and CHEBI\_7754). To ensure we can look up all entities for which activities
are reported, we subclass everything from the BFO 1.1 class MaterialEntity~\cite{Smith2004},
which is a superclass to the CHEMINF Chemical Entity class.

@@

Similarly, properties of small molecules are available from ChEMBL, and as of ChEMBL-RDF 13
these too are exposed in triple format. Like the InChI and InChIKey, these are provided using the
CHEMINF ontology. Here is, for example, the ALogP value (as distributed with ChEMBL and
calculated with ACD/Labs software):

@@

The following SPARQL query returns the 44,681 ChEMBL documents. This is 1 fewer than appears 
in the ChEMBL 13 release notes~\cite{ChEMBL13ReleaseNotes}, as the result does not include the doc\_id -1, as this refers to the 
unpublished dataset, not itself a valid \verb+bibo:Article+.

@@

Notice here the use of \verb+owl:equivalentClass+ since in CHEMINF molecules are defined as
as classes rather than instances.

@@

For example, not all information in the databases has so far been annotated using
standard ontologies. That includes the activity types (``IC50'', ``EC50'', and ``Potency'' in ChEMBL), making it
still impossible to reason or group over similar activity types. For example,
the IC$_{50}$ and K$_i$ end point are closely related, but that is currently not
ontologically expressed. Similarly, measurement units are currently provided as
string literals, making it needlessly hard to compare concentrations in nM and mM.
If we further allow to make some assumptions, we can even search for activities
expressed in $\mu$g/ml with query concentrations in nM. Without introducing machine-readable
semantics, such conversions are less easily automated.

>>>>

For drug discovery, the drug-like compounds themselves are the main topic of study.
ChEMBL contains many different compound types, mostly small molecules,
but also peptides, proteins, antibodies, oligosaccharides, oligonucleotides, and
even cells. ChEMBL-RDF follows the approach used by the CHEMINF ontology,
and encodes drugs as classes, rather than instances, and they subclass
other classes defined by several ontologies for these types. For example, proteins subclass the
Protein class from the PRotein Ontology (PR\_000000001), small
molecules subclass the Chemical Entity class from the CHEMINF
ontology (CHEMINF\_000000) (which is itself equivalent to the Chemical Entity class in ChEBI, CHEBI\_24431), 
and oligosaccharides and oligonucleotides
subclass their respective matches in the CHEBI ontology (CHEBI\_50699
and CHEBI\_7754). To ensure we can look up all entities for which activities
are reported, we subclass everything from the BFO 1.1 class MaterialEntity~\cite{Smith2004},
which is a superclass to the CHEMINF Chemical Entity class.

@@

Similarly, properties of small molecules are available from ChEMBL, and as of ChEMBL-RDF 13
these too are exposed in triple format. Like the InChI and InChIKey, these are provided using the
CHEMINF ontology. Here is, for example, the logP value (as distributed with ChEMBL and
calculated with ACD/Labs software):

@@

The following SPARQL query returns the 44,681 ChEMBL documents. This is one fewer than appears 
in the ChEMBL 13 release notes~\cite{ChEMBL13ReleaseNotes}, as the result does not include the doc\_id -1, as this refers to the 
unpublished dataset, not itself a valid \verb+bibo:Article+.

@@

Notice here the use of \verb+owl:equivalentClass+ since in CHEMINF molecules are defined as
classes rather than instances.

@@

For example, not all information in the databases has so far been annotated using
standard ontologies. That includes the activity types (``IC50'', ``EC50'', and ``Potency'' in ChEMBL), making it
still impossible to reason or group over similar activity types. For example,
the IC$_{50}$ and K$_i$ end points are closely related, but that is currently not
ontologically expressed. Similarly, measurement units are currently provided as
string literals, making it needlessly hard to compare concentrations in nM and mM.
If we further allow to make some assumptions, we can even search for activities
expressed in $\mu$g/ml with query concentrations in nM. Without introducing machine-readable
semantics, such conversions are less easily automated.
